---
title: "Steps Analysis"
author: "Ben Sylve"
date: "5/28/2020"
output: html_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This analysis is intended to provide an overview of personal movement using activity monitoring devices such as Fitbit. The focus is on gaining a better understanding of steps taken throughout the day.

The data were collected during the months of October and November 2012 and includes the number of steps taken in 5 minute intervals each day.

## Load and Preprocess the Data

Load the data set into a table so that analysis can be performed.

```{r}
# Load the dplyr package used to summarize the data
library(dplyr)
# Read in the CSV file of data
activityTable <- read.csv("activity.csv")
# Get the data without the null values
activityWithoutNulls <- activityTable[!is.na(activityTable$steps),]
# Add the data without nulls to a dplyr table
activityWithoutNullsDF <- as_tibble(activityWithoutNulls)

# Get the median and mean steps taken
meanStepsTaken <- as.character(round(mean(stepsByDaySummary$steps),1))
medianStepsTaken <- as.character(round(median(stepsByDaySummary$steps),1))

# Group by date
days <- group_by(activityWithoutNullsDF, date)
# Sum the steps for each date
stepsByDaySummary <- summarise(days, steps = sum(steps))

# Group by interval
intervals <- group_by(activityWithoutNullsDF, interval)
# Sum the number of steps and days
intervalSummary <- summarise(intervals, steps = sum(steps), numOfDays = n_distinct(date))
# Calculate the average
intervalSummary$avgSteps <- intervalSummary$steps / intervalSummary$numOfDays
```

## Overview of Steps Taken Per Day

```{r}
# Produce a histogram of the steps taken per day
hist(stepsByDaySummary$steps,main="Steps Per Day",xlab="Number of Steps",col="blue")
```

The mean number of steps per day was `r meanStepsTaken` while the median was `r medianStepsTaken`.

## Average Daily Activity Pattern

```{r}
# Produce a line chart of the average number of steps taken per interval
plot(intervalSummary$interval, intervalSummary$avgSteps, type="l", xlab="Interval", ylab="Avg. Number of Steps", main="Average Number of Steps Per Interval")
```
