---
title: "Steps Analysis"
author: "Ben Sylve"
date: "5/28/2020"
output: html_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This analysis is intended to provide an overview of personal movement using activity monitoring devices such as Fitbit. The focus is on gaining a better understanding of steps taken throughout the day.

The data were collected during the months of October and November 2012 and includes the number of steps taken in 5 minute intervals each day.

## Load and Preprocess the Data

Load the data set into a table so that analysis can be performed.

```{r}
# Load the dplyr package used to summarize the data
library(dplyr)
library(ggplot2)

# Read the csv data and store it as a tables
activityTable <- read.csv("activity.csv")
activityTableDF <- as_tibble(activityTable)
activityWithoutNulls <- activityTable[!is.na(activityTable$steps),]
activityWithoutNullsDF <- as_tibble(activityWithoutNulls)

# Total the steps by date
days <- group_by(activityWithoutNullsDF, date)
stepsByDaySummary <- summarise(days, steps = sum(steps))

# Get the mean and median steps per day
meanStepsTaken <- as.character(round(mean(stepsByDaySummary$steps),1))
medianStepsTaken <- as.character(round(median(stepsByDaySummary$steps),1))

# Get the average number of steps per interval
intervals <- group_by(activityWithoutNullsDF, interval)
intervalSummary <- summarise(intervals, steps = sum(steps), numOfDays = n_distinct(date))
intervalSummary$avgSteps <- intervalSummary$steps / intervalSummary$numOfDays

# Get the interval with the highest average number of steps
invervalWithMaxStepsOnAverage <- intervalSummary[intervalSummary$avgSteps == max(intervalSummary$avgSteps),]$interval

# Get the number of records with null steps recorded
numberOfRecordsWithNoData <- sum(is.na(activityTable$steps))

# Default the steps with missing values substituted to the steps

# Join the data with nulls to the interval summary based on the summary by interval
activityTableWithIntervalAvgsDF <- left_join(activityTableDF, intervalSummary, by = "interval")
# Add a column to store the number of steps with the substituted value for nulls
activityTableWithIntervalAvgsDF$stepsNullsSubstituted <- activityTableWithIntervalAvgsDF$steps.x
# Get the rows with nulls
index <- is.na(activityTableWithIntervalAvgsDF$stepsNullsSubstituted)
# Set the nulls to the average for that interval
activityTableWithIntervalAvgsDF$stepsNullsSubstituted[index] <- activityTableWithIntervalAvgsDF$avgSteps
# Add a field to determine if the day is a weekday or weekend
activityTableWithIntervalAvgsDF$dayOfWeek <- weekdays(as.Date(activityTableWithIntervalAvgsDF$date))
activityTableWithIntervalAvgsDF$dayType <- "Weekday"
activityTableWeekendIndex <- activityTableWithIntervalAvgsDF$dayOfWeek == "Saturday" | activityTableWithIntervalAvgsDF$dayOfWeek == "Sunday"
activityTableWithIntervalAvgsDF$dayType[activityTableWeekendIndex] <- "Weekend"

# Add weekend summary
summaryByDayType <- activityTableWithIntervalAvgsDF %>% group_by(dayType, interval) %>% summarize(avgStepsPerDayType = mean(stepsNullsSubstituted, na.rm=TRUE))

# Make an equivalent table with the subsitute value
newActivityTable <- select(activityTableWithIntervalAvgsDF, stepsNullsSubstituted, date, interval)
newActivityTableDF <- as_tibble(newActivityTable)

# Total the steps by date
newDays <- group_by(newActivityTableDF, date)
newStepsByDaySummary <- summarise(newDays, steps=sum(stepsNullsSubstituted))

# Get the mean and median steps per day
newMeanStepsTaken <- as.character(round(mean(newStepsByDaySummary$steps),1))
newMedianStepsTaken <- as.character(round(median(newStepsByDaySummary$steps),1))

newActivityTableDF$dayOfWeek <- weekdays(as.Date(newActivityTableDF$date))
newActivityTableDF$dayType <- "Weekday"
weekendIndex <- newActivityTableDF$dayOfWeek == "Saturday" | newActivityTableDF$dayOfWeek == "Sunday"
newActivityTableDF$dayType[weekendIndex] <- "Weekend"

weekendIntervalGroup <- group_by(newActivityTableDF[newActivityTableDF$dayType=="Weekend",], interval)
weekendIntervalSummary <- summarise(weekendIntervalGroup, steps = sum(stepsNullsSubstituted), numOfDays = n_distinct(date))
weekendIntervalSummary$avgSteps <- weekendIntervalSummary$steps / weekendIntervalSummary$numOfDays

weekdayIntervalGroup <- group_by(newActivityTableDF[newActivityTableDF$dayType=="Weekday",], interval)
weekdayIntervalSummary <- summarise(weekdayIntervalGroup, steps = sum(stepsNullsSubstituted), numOfDays = n_distinct(date))
weekdayIntervalSummary$avgSteps <- weekdayIntervalSummary$steps / weekdayIntervalSummary$numOfDays
#http://zevross.com/blog/2019/04/02/easy-multi-panel-plots-in-r-using-facet_wrap-and-facet_grid-from-ggplot2/
```

## Total Number of Steps Taken Per Day

```{r}
# Display a histogram of the number of steps per day
hist(stepsByDaySummary$steps,main="Steps Per Day",xlab="Number of Steps",col="blue")
```

The mean number of steps per day was `r meanStepsTaken` while the median was `r medianStepsTaken`.

## Average Daily Activity Pattern

The interval with the highest average steps is `r invervalWithMaxStepsOnAverage`

```{r}
# Display a line chart of the average number of steps per interval
plot(intervalSummary$interval, intervalSummary$avgSteps, type="l", xlab="Interval", ylab="Avg. Number of Steps", main="Average Number of Steps Per Interval")
```

## Imputing Missing Values

The data set contains `r numberOfRecordsWithNoData` missing values.

```{r}
# Display a histogram of the number of steps per day
hist(newStepsByDaySummary$steps,main="Steps Per Day",xlab="Number of Steps",col="blue")
```

The mean number of steps per day was `r newMeanStepsTaken` while the median was `r newMedianStepsTaken`.

There is not a major difference with the imputed values

## Comparing Activity Patters of Weekdays to Weekends

```{r}
#plot(weekendIntervalSummary$interval, weekendIntervalSummary$avgSteps, type="l", xlab="Interval", ylab="Avg. Number of Steps", main="Average Number of Steps Per Interval")
#plot(weekdayIntervalSummary$interval, weekdayIntervalSummary$avgSteps, type="l", xlab="Interval", ylab="Avg. Number of Steps", main="Average Number of Steps Per Interval")
ggplot(data=summaryByDayType, aes(x=interval, y=avgStepsPerDayType, group=1)) + geom_line() + geom_point() + facet_wrap(~ dayType) + labs(y="Average Steps per Day", x="Interval")
```